---
title: "Class12"
author: "Sabrina Wu"
date: "2024-11-07"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Section 1. Propoertion of G/G in a population

Downloaded a CSV file from Ensemble < https://useast.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39894595-39895595;v=rs8067378;vdb=variation;vf=959672880#373531_tablePanel>

Here we read this CSV file

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
```
```{r}
table(mxl$Genotype..forward.strand.)
```
```{r}
table(mxl$Genotype..forward.strand.)/nrow(mxl)*100
```
Now let's look at a different population. I picked the GBR.

```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
```
Find proportion of G|G

```{r}
round(table(gbr$Genotype..forward.strand.)/nrow(gbr)*100,2)
```
This variant that is associated with childhood asthma is more frequent in the GBR population than the MKL population.

Lets now idg into this further. 

##Section 4: Population Scale Analysis

One sample is obviously not enough to know what is happening in a population. You are 
interested in assessing genetic differences on a population scale. 


>Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes. Hint: The read.table(), summary() and boxplot() functions will likely be useful here. There is an example R script online to be used ONLY if you are struggling in vein. Note that you can find the medium value from saving the output of the boxplot() function to an R object and examining this object. There is also the medium() and summary() function that you can use to check your 
understanding.


How many samples do we have?
```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```
```{r}
table(expr$geno)
```
```{r}
summary(expr)
```
```{r}
library(dplyr)

medians <- expr %>%
  group_by(geno) %>%
  summarize(median(exp, na.rm=TRUE))
medians
```

There are 462 samples total where 108 corresponding to A/A, 233 for A/G, and 121 for G/G. The median expression level of A/A is 31.24, A/G is 25.06, and G/G is 20.07. 



>Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3? Hint: An example boxplot is provided overleaf – yours does not need to be as polished as this one

```{r}
library(ggplot2)
```

Let's make a boxplot

```{r}
boxplot <- ggplot(expr)+ aes(x=geno, y=exp, fill=geno) +
  geom_boxplot(notch=TRUE)+
  geom_jitter(alpha=0.2)
boxplot
```
There are more A/A expression than the other two types of genotypes since the median, and quartiles are higher on the boxplot. SNP does effect the expression of ORMDL3. 






