---
title: "Class10: Halloween Mini-Project"
author: "Sabrina Wu (A16731683)"
format: pdf
---

## Importing Candy data

Download dataset

```{r}
candy_file <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv"

candy = read.csv(candy_file, row.names=1)
head(candy)
```
>Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```
There are 85 different types of candies. 

>Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```
There are 38 types of fruity candy. 

## What is your favorite candy?

```{r}
candy["Twix", ]$winpercent

```
>Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy["Nestle Crunch",]$winpercent
```
Nestle Crunch winpercent is 66.5%

>Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```
Kit Kat winpercent ifs 76.7%.

>Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars","winpercent"]
```
The winpercent of Tootsie Roll Snack Bars is 49.6%

```{r}
library(dplyr)

candy |>
  filter(rownames(candy)=="Haribo Happy Cola") |>
  select(winpercent)
```
>Q. Find fruity candy with a winpercent above 50%

```{r}
candy |>
  filter(winpercent > 50) |>
  filter(fruity ==1)
```
```{r}
top.candy <- candy[candy$winpercent >50,]
top.candy[top.candy$fruity ==1,]
#this is more complicated than the one above
```

Install Skimer package. To get a quick insight into a new dataset some folks like using the skimer package and its `skim()` function. 

```{r}
#install.packages("skimer") or just go through package
skimr::skim(candy)
```
>Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

The `winpercent` column/variable is around 100 times the other columns since the other columns are all below 1. 
(Need to scale the data before doing any analysis like PCA etc.)

>Q7. What do you think a zero and one represent for the candy$chocolate column?

A zero and one represent if the candy is classified as a choloate candy or not. 

>Q8. Plot a histogram of winpercent values

We can do this a few ways, e.g. the "base" R `hist()` function or with `ggplot()`
```{r}
hist(candy$winpercent, breaks = 8)
```
```{r}
library("ggplot2")

ggplot(candy)+
  aes(winpercent)+
  geom_histogram(binwidth =8) +
  theme_bw()
```


>Q9. Is the distribution of winpercent values symmetrical?

No, the winpercent seems to be skewed right, where theres more value to the left. 

>Q10. Is the center of the distribution above or below 50%?

```{r}
summary(candy$winpercent)
```
THe median is below 50% and the median is above 50. The median is a better comparision as the data is skewed. 

>Q11. On average is chocolate candy higher or lower ranked t
than fruit candy?

```{r}
fruit.candy <- candy |>
  filter(fruity==1)
chocolate.candy <- candy |>
  filter(chocolate ==1)

summary(fruit.candy$winpercent)
summary(chocolate.candy$winpercent)

#this is easier to read it compared to the one below it
  
```
```{r}
summary(candy[as.logical(candy$fruity),]$winpercent)
summary(candy[as.logical(candy$chocolate),]$winpercent)
```
Chocolate candy is ranked above fruity candy as all the summary value (min, 1st quartile, median, mean, 3rd quartile, and max) are all greater than the one for fruity. 

>Q12. Is this difference statistically significant?

```{r}
t.test(chocolate.candy$winpercent, fruit.candy$winpercent)
```
P-value is extremely low so there is a significant difference.

## Overall Candy Rankings

>Q13. What are the five least liked candy types in this set?

```{r}
play <- c("d","a","c")
sort(play)
order(play)

play[order(play)]
#use play instead of order bc want to know what the candy is and not just the top winpercent value
```
```{r}
head(candy[order(candy$winpercent),])
```
Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, Jawbusters, and Root Beer Barrels are the least liked candies. 


>Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
candy |>
  arrange(winpercent) |>
  tail(5)
```
```{r}
head(candy[order(candy$winpercent, decreasing=T),])
```

Reese, twix, kitkat, and snickers are the top 5 all time favorite candies. 

>Q15. Make a first barplot of candy ranking based on winpercent values.

Let's do a barplot of winpercent values
```{r}
ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()
```


>Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),-winpercent)) +
  geom_col()
```

Adding some useful color

```{r}
ggplot(candy) + 
  aes(x=winpercent, 
      y=reorder(rownames(candy),winpercent), fill=chocolate) +
  geom_col()
```

I want a more custom color scheme where I can see both chocolate an bar and fruity etc. all from the one plot. To do this we can roll our own color vector...


```{r}
#this is the place holder color vector
my_cols=rep("black",nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"

```
```{r}
#Use blue for your favorite candy!
my_cols[rownames(candy)=="Nestle Crunch"] <-"blue"

```

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

>Q17. What is the worst ranked chocolate candy?

The worst ranked chocolate candy is sixlets. 

>Q18. What is the best ranked fruity candy?

Starburst is the best ranked fruity candy. 

## 4. Taking a look at pricepercent

Plot of winpercent vs pricepercent to see what would be the best candy to buy...

Overwriting the fruity from pink to red so it is more visible. 
```{r}
my_cols[as.logical(candy$fruity)] <- "red"
```


```{r}
ggplot(candy)+
  aes(x=winpercent, y=pricepercent)+
  geom_point(col=my_cols)
```

Add labels
```{r}
ggplot(candy)+
  aes(x=winpercent, y=pricepercent, label=rownames(candy))+
  geom_point(col=my_cols)+
  geom_text()
#this is hard to read bc of the overlaps, use ggrepel
```
```{r}
library(ggrepel)

ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```


>Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

The Reese's Miniature is the best bang for your buck. 


>Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?


```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

The most expensive is Nik L Nip, Ring pop, Nestle Smarties, Milky Way, Herskey's. Nik Nip is the least popular.'

> Q21. Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment()

```{r}
ggplot(candy) + 
  aes(pricepercent, reorder(rownames(candy),pricepercent)) +
  geom_col()
```
```{r}
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()
```
A lot of the the candies are the same price points

## 5. Exploring the correlation structure

```{r}
library(corrplot)
```
```{r}
cij <- cor(candy)
corrplot(cij, diag =F)
```

>Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Fruity and chocolate are strongly anti-correlated. 

>Q23. Similarly, what two variables are most positively correlated?

Chocolate and winpercent are the most positively correlated.
```{r}
cij
```


## 6. Principal Componenent Analysis

```{r}
pca <- prcomp(candy, scale = TRUE)
summary(pca)
```
```{r}
plot(pca$x[,1:2])
```

```{r}
plot(pca$x[,1], pca$x[,2], col=my_cols, pch=16)
```

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

Making labels
```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```
Make interactive with plotly package
```{r}
#library(plotly)
#ggplotly(p)
```
```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

How do the original variables(columns) contribute to the new PCs. I will look at PC1 here

This is the same plot as before but reorder and using different ones. 

```{r}
loadings <- as.data.frame(pca$rotation)

ggplot(loadings) +
  aes(PC1,reorder(rownames(loadings),PC1),PC1, fill=PC1) +
  geom_col()
```

>Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Fruity are strongly by PC1 in the positive and negative is chocolate. 



